Semantic authoring in a flexiformal context --- Bulk annotation of rigorous documents

### 2025-05-04 doc/cicm25-sysdesc/sysdesc.tex
better title
work on paper

\documentclass{llncs}
\usepackage[show]{ed}
\usepackage{wrapfig}
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{stex-logo}
\usepackage{paralist}
%\usepackage{lststex} debug that the gray is not solid
\usepackage{listings}
\definecolor{codegray}{rgb}{0.9,0.9,0.9}
\lstset{basicstyle=\sf,columns=fullflexible,backgroundcolor = \color{codegray}}
\lstset{numberstyle=\tiny}
\lstset{language={[LaTeX]TeX}}
\usepackage[style=alphabetic,hyperref=auto,defernumbers=true,backend=bibtex,firstinits=true,maxbibnames=9,maxcitenames=3,isbn=false]{biblatex}
\addbibresource{kwarcpubs.bib}
\addbibresource{extpubs.bib}
\addbibresource{kwarccrossrefs.bib}
\addbibresource{extcrossrefs.bib}
\usepackage[noabbrev]{cleveref}

@@

\newcommand\ALeA{\textsf{ALeA}\xspace}
\newcommand\SMGloM{\textsf{SMGloM}\xspace}
\newcommand\snify{\textsf{snify}\xspace}
\def\llangle{\langle\kern-.2em\langle}
\def\rrangle{\rangle\kern-.2em\rangle}

\title{
    Semantic Authoring in the \sTeX/\ALeA World
    \ednote{Find better title. Was ``Bulk Semantic Annotation with a Partially Known Knowledge Base''
}}
\author{Michael Kohlhase, Jan Frederik Schaefer}
\institute{Computer Science, FAU Erlangen N\"urnberg, Germany}
\begin{document}
\maketitle
\begin{abstract}
  Semantic annotations can be utilized for learning support services in technical
  documents. The resulting active documents support more direct and personalized
  interaction with the knowledge conveyed, but require a large investment in the authoring
  process.

  In this paper we discuss the nature of the semantic authoring problem and argue that
  while sharing aspects of classical, informal authoring and formal authoring, it is a
  problem of its own that needs special editing support facilities. We present a simple
  utility that supports the process of annotating technical terms with semantic references
  to a flexiformal content commons that enhances annotation productivity by almost an
  order of magnitude.
\end{abstract}

\section{Introduction}
Arguably, dealing with large document collections is one of the key factors in the knowledge-driven society and economy.
There are currently two main contenders for machine support in this area.
Symbolic/logic-based technologies and sub-symbolic/machine-learning-based AI, e.g.\ via LLMs or chatbots.
They have complementary strengths and challenges:
ML-based approaches can be trained on all data of the Internet, but face challenges in precision and explainability. 
Symbolic technologies, on the other hand, offer precision and explainability out of the box, but face scalability challenges
-- in particular because of the cost of semantic authoring\ednote{
    \emph{semantic authoring} already used in context of semantic web (RDFa).
    Tools exist and it doesn't have exactly the same challenges as we face, though there is some overlap.
    We have a module system and need more dependency management, something else?
    This has to be described somewhere.

    Related work: ``User Interfaces for Semantic Authoring of Textual Content: A Systematic Literature Review.''
}, which we discuss in this paper.

As a case study, we will use adaptive learning assistants for STEM subjects in higher education, concretely the \ALeA system \cite{BerBetChu:lssmkm23}.
We contend that machine-learning based technologies are largely unsuitable for this task -- at least until they reach the
precision and explainability required by the ethics of teaching --
while the scope of specific domains is just narrow enough that symbolic approaches are viable.
Conventional (\LaTeX-based) lecture materials can essentially be directly imported into \ALeA,
but its ability to generate learner-adaptive learning objects -- instrumented with learning support interactions --
depends on semantic annotations.
These can be embedded \LaTeX\ documents using the \sTeX package~\cite{MueKo:sdstex22,sTeX:github:on}.
In this paper, we will mostly focus on annotating terms with semantic concept they refer to,
but \sTeX supports a much wider range of annotations, such as annotating problems with the learning objectives
or annotating the defining phrase (definiens) in concept definitions,
and can, in principle, even be used for full formalization.\ednote{True?}
As \sTeX supports the full spectrum from informal (un-annotated) to fully formalized content,
we also call \sTeX content \textbf{flexiformal}.

The \textbf{\sTeX/\ALeA content commons} consists of \SMGloM~\cite{GinIanJuc:spsttom16},
the domain model for semantic annotations, as well as re-usable lecture materials such as
lecture slides and problems -- mostly from the area of computer science.  The sheer size
of the commons is one of the key challenges for authors: \SMGloM contains definitions of
$X$\ednote{TODO} STEM concepts in multiple languages, created and maintained by over 30
contributors.  Additionally, there are roughly 4900\ednote{ Includes 440 from UPS, which
  seems un-annotated...  Excludes talks as well...  Other issues?  Command (tex
  comment): % grep -R --include='*.tex' -F -e '\begin{frame}' | wc -l
} semantically annotated lecture slides and 4000 problems.

@@

% % OLD INTRO
% Arguably, dealing with large document collections is one of the key factors in the
% knowledge-driven society and economy. There are currently two main contenders for machine
% support in this area. Symbolic/logic-based technologies and sub-symbolic,
% machine-learning-based AI, e.g.\ via LLMs or chatbots. They have complementary strengths
% and challenges: Symbolic technologies offer precision and explainability out of the box,
% but face scalability challenges because the prerequisite background knowledge has to be
% (manually) formalized. ML-based approaches can be trained on all data of the Internet, but
% face challenges in precision and explainability. In \cite{Ranta:atcp17},
% Aarne Ranta matches these profiles to the notion of \textbf{producer tasks} -- i.e.\ tasks
% where precision is key, but limited coverage ($10^{3\pm1}$ concepts) is OK;
% e.g.\ for multi-language/variant manuals for very
% expensive machines -- and \textbf{consumer tasks}, where coverage is key and
% precision secondary; e.g.\ for consumer-grade machine translation like \textsf{Google translate}.
% 
% In this paper, we address tools that help attack the coverage problem in symbolic/semantic
% approaches in practice. The producer task we use as a case study is that of adaptive
% learning assistants for STEM subjects, concretely the \ALeA system \cite{BerBetChu:lssmkm23}.
% \ALeA uses \sTeX \cite{MueKo:sdstex22,sTeX:github:on} -- a
% variant of {\LaTeX} that allows to embed semantic annotations -- for knowledge
% representation, and generates learner-adaptive learning objects instrumented with
% learning-support interactions from that. The \textbf{\sTeX/\ALeA content commons}
% \ednote{MK@MK: continue to describe the size and distribution, and
%   authorship}\ednote{explain the \textbf{domain model} (the SMGloM) and the
%   \textbf{formulation model} (cf. \cite{BerBetChu:lssmkm23}) in the commons and give their
%   relative sizes.}
% 
% We contend that while education -- especially higher education -- as a whole is a
% wide-coverage task for society education of individual learners in specific domains --
% especially, if we want to tailor the educational offerings to said individual/domain -- is
% a producer task, at least until ML-based methods reach the precision and explainability
% required by the ethics of teaching.

@@

Before we discuss semantic authoring in general, let us introduce a concrete example from
the \sTeX/\ALeA world. We will use it to sharpen our intuitions of the problems and issues
involved and to ground the solution we propose.

@@

% The main practical problem of annotating presentational {\LaTeX} course materials into
% semantic documents that carry enough information to support meaningful learning support
% services is to add semantic references to technical terms from the domain jargon.

Concretely this is about converting the {\LaTeX} string\ednote{MK: There is/was a discrepancy
  in the last line of the example; I tried to fix it, but failed}
\begin{lstlisting}[numbers=left,firstnumber=3,
caption=The unannotated {\LaTeX} sources of \cref{fig:lo},label=lst:los]
\item \blue{Note:} Depth-limited minimax requires an evaluation for every
cut-off state $s$. If $s$ is terminal, we use its utility, and otherwise
an estimate.
\end{lstlisting}
into the partially annotated item:
\begin{lstlisting}[morekeywords={sr,importmodule},numbers=left,
caption=Annotating ``terminal'' and ``utility'' in \cref{lst:los},label=lst:losa]
\importmodule[smglom/search]{mod?search-problem}
[...]
\item \blue{Note:} Depth-limited minimax requires an evaluation for every
cut-off state $s$. If $s$ is \sr{goal state}{terminal}, we use its \sn{utility},
and otherwise an estimate.
\end{lstlisting}
In the latter, the author has the word ``terminal'' with the concept it refers to with 
via the \sTeX macro
\lstinline[mathescape]|\sr{$\llangle URI \rrangle$}{$\llangle verbalization\rrangle$}|.

\begin{newpart}{MK: more definitions}
  In \sTeX, objects and concepts are represented by \textbf{symbols}, which are identified
globally and unambiguously by a URI that includes a \textbf{namespace}, a \textbf{module
  name}, and a \textbf{symbol name}. Every symbol can have one or more
\textbf{verbalizations}: words that of the domain jargon used to refer to the concept or
object. Verbalizations can vary between between subdomains and communities of practice. In
our example, the verbalization ``terminal [state]'' was annotated with the symbol with
name \lstinline|goal state|. It comes from the module with name \lstinline|search-problem|
in the namespace \lstinline|smglom/search/mod| imported via the \lstinline|\importmodule|
macro in line 1. As the symbol name \lstinline|goal state| is unique in the imports here
it suffices as a relative URI in the \lstinline|\sr| annotation.

If the symbol name and the desired verbalization coincide (as they often do in English),
the short-hand macro \lstinline|\sn| can be used (e.g. as \lstinline|\sn{utility}| in
\cref{lst:losa}).
\end{newpart}

@@

\begin{figure}[ht]\centering
  \fbox{\includegraphics[width=12cm]{../img/search-problem.en.pdf}}
\begin{lstlisting}[morekeywords={definame,symdecl,definiendum},numbers=left]
\begin{smodule}[title=Search Problem]{search-problem}
[... some imports ...]
\symdecl*{goal state}
\begin{sdefinition}
  [...] Certain \sns{state} are [...] \definame[post=s]{goal state} [...]
  (also called \definiendum{goal state}{terminal states}).
\end{sdefinition}
\end{smodule}
\end{lstlisting}
  \caption{Simplified definition of ``goal state'' from the domain model}\label{fig:state-space}
\end{figure}

\begin{newpart}{MK: trying to make some general remark about annotation before we go into
    semantic authoring}
\subsection{Annotation Workflows and Requirements}

To annotate the word ``terminal'' in the document, the author has to be aware that it is a
technical term, that the module from \cref{fig:state-space} exists, know the symbol URL,
and manage redundancy of the imports -- i.e. only adding the \lstinline|\importmodule|
directive if it is not already (recursively) implied and possibly removing directives that
become redundant by the new one.  At ca 150,000 words in a semester's worth of lecture
notes and ca. 10--15\% of technical terms a rather daunting task.

@@

The most important resource for annotators is a verbalization to symbol URI mapping -- we
call it the \textbf{annotation catalog} -- either implicitly in the mind of the annotator
or explicitly as a technical artifact, which allows to
\begin{inparaenum}[\em i\rm)]
\item identify the known technical terms,
\item to map them to the corresponding symbols, and 
\item manage the necessary import directives.
\end{inparaenum}
This catalog can be harvested from the \textbf{formal declarations}
(\lstinline|\definiendum|, \lstinline|\begin|/\lstinline|\end{smodule}| and
\lstinline|\importmodule|, etc.) of the domain model and existing documents in the
semantic corpus.

@@

But still, $L$-annotation catalogs are usually inadequate for $L$-annotation tasks in
practice, as
\begin{enumerate}[\em i\rm)]
\item $L$-verbalizations and symbol names do not often coincide and
\item the domain model may not even have $L$-modules -- especially for the ``smaller''
  languages
\end{enumerate}
 In this situation, we usually have to invest in
establishing a suitable catalog e.g. by providing $L$-translations (or at least dictionary
information) in the domain model or delineating modules and annotating definienda that the
domain model does not cover in the documents to be annotated.
\end{newpart}

@@

\section{Tool Support for Semantic Authoring in \sTeX}\label{sec:tools}
Semantic authoring can ``inherit'' tool support from both informal and formal authoring.
In the case of \sTeX, we can re-use spell checkers for traditional informal authoring.
Additionally, there is a custom VSCode IDE plugin \cite{sTeX-IDE:git} that analyzes
annotations on the fly, displays the underlying reference semantics, reports errors and
redundancies, and even offers a concept search interface.  But such features only support
existing annotations and are, therefore, of limited use when creating new annotations,
which arguably is the most time consuming aspects of semantic authoring.  \ednote{also:
  SPARQL API}

@@

Efficient support for the annotation of technical terms in natural language
has been elusive so far.
\ednote{
    Move Hutter from conclusion here?
    Also: Plivelic}

\section{The \snify System}\label{sec:snify}

The \snify utility (from ``\texttt{\textbackslash sn}-ify'' for adding term references;
see \cite{stextools:git}) is a simple command-line tool that creates a database of
symbol-verbalization pairs, analyzes \sTeX source files, and then steps through all word
occurrences in the document that match a verbalization in the database.  Matching is based
on word stems using off-the-shelf stemmers, which allows us to e.g.\ match ``utilities''
with ``utility''.  For each matched word, the user is presented with an annotation choice
and interactions that allow to fine-tune the wordwise annotation
workflow. \Cref{fig:snify} shows a typical situation.

@@

\section{Practical Evaluation}\label{sec:eval}

In our experience, \snify's step-through workflow for annotating term references for symbols
from the domain model is almost an order of magnitude more efficient than writing the
annotations and imports by hand, even for annotators who are familiar with the domain
model. For annotators unfamiliar with the domain model, the unassisted annotation task is
almost infeasible, and the average unfamiliarity naturally grows with the domain
model. The symbol disambiguation process -- on average a word induces 4.1 choices\ednote{
    more details? median is 3, for existing annotations, the average is 2.6 and the median 2...
} -- is still manageable,
requires considerable concentration and domain knowledge, but little knowledge of the
domain model flexiformalization.

@@

For less complete domain models we have to skip too many terms that should ultimately be
annotated and annotation efficiency suffers. This is currently the case for all
non-English languages in the \sTeX corpus we work with. Coverage in German is only about
half of that of English and we can already see the practical effects. We have also
experimented with a Slovene introductory math book, and it seems clear that apart from
having to introduce a Slovene stemmer\ednote{MK@FS: there is one for python at
  \url{https://repo.ijs.si/pboskoski/slo_stemmer}}, we would need to manually annotate all
definienda in the book before we can harvest the symbol/verbalization pairs which are a
prerequisite for annotation.

One practical problem with \snify that remains is that the system only harvests
symbol/verbalization pairs from \sTeX files on the local file system. This means that, for
effective annotation, users (have to) download the whole content commons and keep it
updated. Unfortunately, the current MMT-based MathHub knowledge management
API\ednote{intro this above} does not supply verbalization information. The iMMT-based
API~\cite{iMMT:on} currently under development will, so \snify will be able to request the
pairs for the remaining (non-local) content commons. If a pair is chosen in an annotation,
\snify can then automatically download the corresponding archive so that the annotated
file can be formatted with pdflatex or rusTeX locally.

@@

In an earlier attempt to support semantic annotation in IDEs, we tried named-entity
recognition (NER) for classification of ``likely annotation candidate words''
\cite{hutterer:msc23}, however this classification was not precise enough in the
distinction in ``technical terms'' and ordinary English noun phrases and named entities --
the relevant task for annotation, and so made it impractical, especially, since it could
not enter the actual annotations and import directives automatically. But maybe using the
NER-based approach, after the \snify annotation process might change the trade-offs
involved. In any case, the overall workflow suggested by the NER-based approach --
especially after the verbalizations covered by the existing domain model\ednote{introduce
  in the definition above that the domain model is all definitions, not only in the
  smglom?} -- is geared towards adding modules and symbols -- the ones discovered by NER
-- to the domain model.

>>>>

\documentclass{llncs}
\usepackage[show]{ed}
\usepackage{wrapfig}
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{stex-logo}
\usepackage{paralist}
\usepackage{wrapfig}
%\usepackage{lststex} debug that the gray is not solid
\usepackage{listings}
\definecolor{codegray}{rgb}{0.9,0.9,0.9}
\lstset{basicstyle=\sf,columns=fullflexible,backgroundcolor = \color{codegray}}
\lstset{numberstyle=\tiny}
\lstset{language={[LaTeX]TeX}}
\usepackage[style=alphabetic,hyperref=auto,defernumbers=true,backend=bibtex,firstinits=true,maxbibnames=9,maxcitenames=3,isbn=false]{biblatex}
\addbibresource{kwarcpubs.bib}
\addbibresource{extpubs.bib}
\addbibresource{kwarccrossrefs.bib}
\addbibresource{extcrossrefs.bib}
\usepackage[noabbrev]{cleveref}

@@

\newcommand\ALeA{\textsf{ALeA}\xspace}
\newcommand\SMGloM{\textsf{SMGloM}\xspace}
\newcommand\snify{\textsf{snify}\xspace}
\newcommand\defianno{\textsf{defianno}\xspace}
\def\llangle{\langle\kern-.2em\langle}
\def\rrangle{\rangle\kern-.2em\rangle}

\title{
    % Semantic Authoring in the \sTeX/\ALeA World
    % \ednote{Find better title. Was ``Bulk Semantic Annotation with a Partially Known Knowledge Base''}
Semantic authoring in a flexiformal context --- Bulk annotation of rigorous documents
}
\author{Michael Kohlhase, Jan Frederik Schaefer}
\institute{Computer Science, FAU Erlangen N\"urnberg, Germany}
\begin{document}
\maketitle
\begin{abstract}
%   Semantic annotations can be utilized for learning support services in technical
%   documents. The resulting active documents support more direct and personalized
%   interaction with the knowledge conveyed, but require a large investment in the authoring
%   process.
% 
%   In this paper we discuss the nature of the semantic authoring problem and argue that
%   while sharing aspects of classical, informal authoring and formal authoring, it is a
%   problem of its own that needs special editing support facilities. We present a simple
%   utility that supports the process of annotating technical terms with semantic references
%   to a flexiformal content commons that enhances annotation productivity by almost an
%   order of magnitude.

    Semantic annotations enable services that make the knowledge
    in documents more accessible.
    In the case of \ALeA, a learning assistant we use as a case study,
    they enable, for example, guided tours and practice problems
    tailored to the learner's competencies.
    However, the cost of semantic authoring is substantial
    and tool support is limited.

    In this paper we discuss the nature of the semantic authoring
    problem and argue that, while sharing aspects of classical
    (informal) authoring and formal authoring,
    it is a problem of its own that needs special editing support facilities.
    We present a simple utility that supports the process of annotating
    technical terms with semantic references to a flexiformal content commons
    that enhances annotation productivity by almost an order of magnitude.
\end{abstract}

\section{Introduction}
Arguably, dealing with large document collections is one of the key factors in the knowledge-driven society and economy.
There are currently two main contenders for machine support in this area.
Symbolic/logic-based technologies and sub-symbolic/machine-learning-based AI, e.g.\ via LLMs or chatbots.
They have complementary strengths and challenges:
ML-based approaches can be trained on all data of the Internet, but face challenges in precision and explainability. 
Symbolic technologies, on the other hand, offer precision and explainability out of the box, but face scalability challenges
-- in particular because of the cost of semantic authoring, which we discuss in this paper.

As a case study, we will use adaptive learning assistants for STEM subjects in higher education, concretely the \ALeA system \cite{BerBetChu:lssmkm23}.
We contend that machine-learning based technologies are largely unsuitable for this task -- at least until they reach the
precision and explainability required by the ethics of teaching --
while the scope of specific domains is just narrow enough that symbolic approaches are viable.
Conventional (\LaTeX-based) lecture materials can essentially be directly imported into \ALeA,
but its ability to generate learner-adaptive learning objects -- instrumented with learning support interactions --
depends on semantic annotations.
These can be embedded \LaTeX\ documents using the \sTeX package~\cite{MueKo:sdstex22,sTeX:github:on}.
In this paper, we will mostly focus on annotating terms with the semantic concepts they refer to,
but \sTeX supports a much wider range of annotations, such as annotating problems with the learning objectives
or annotating the defining phrase (definiens) in concept definitions,
and can, in principle, even be used for full formalization.\ednote{True?}
As \sTeX supports the full spectrum from informal (un-annotated) to fully formalized content,
we say that \sTeX content is \textbf{flexiformal}.

The \textbf{\sTeX/\ALeA content commons} consists of \SMGloM~\cite{GinIanJuc:spsttom16},
a glossary acting as the domain model for semantic annotations,
as well as re-usable lecture materials such as lecture slides and problems.
% -- mostly from the area of computer science.
The sheer size
of the commons is one of the key challenges for authors: \SMGloM contains definitions of over
5400 concepts in multiple languages, created and maintained by over 30
contributors.  Additionally, there are roughly 4900
semantically annotated lecture slides and 4000 problems
-- mostly from the area of computer science.
% grep -R --include='*.tex' -F -e '\begin{frame}' | wc -l

\ednote{
    \textbf{Related work: semantic web authoring}
    \emph{semantic authoring} already used in context of semantic web (RDFa).
    Tools exist and it doesn't have exactly the same challenges as we face, though there is some overlap.
    We have a module system and need more dependency management, something else?
    This has to be described somewhere.

    Related work: ``User Interfaces for Semantic Authoring of Textual Content: A Systematic Literature Review.''
}

@@

@@

Before we discuss semantic authoring in general, let us introduce a concrete example from
the \sTeX/\ALeA world. We will use it to sharpen our intuitions of the challenges
involved and to ground the solution we propose.

@@

Concretely this is about converting the {\LaTeX} string
    (for conciseness completely unannotated)
    \ednote{MK: There is/was a discrepancy
  in the last line of the example; I tried to fix it, but failed\\
    FS: Is it fixed now? I don't see any discrepancy.}
\begin{lstlisting}[numbers=left,firstnumber=3,
caption=The unannotated {\LaTeX} sources of \cref{fig:lo},label=lst:los]
\item \blue{Note:} Depth-limited minimax requires an evaluation for every
cut-off state $s$. If $s$ is terminal, we use its utility,
and otherwise an estimate.
\end{lstlisting}
into the partially annotated item:
\begin{lstlisting}[morekeywords={sr,importmodule},numbers=left,
caption=Annotating ``terminal'' in \cref{lst:los},label=lst:losa]
\importmodule[smglom/search]{mod?search-problem}
[...]
\item \blue{Note:} Depth-limited minimax requires an evaluation for every
cut-off state $s$. If $s$ is \sr{goal state}{terminal}, we use its utility,
and otherwise an estimate.
\end{lstlisting}
In the latter, the author has the word ``terminal'' with the concept it refers to with 
via the \sTeX macro
\lstinline[mathescape]|\sr{$\llangle URI \rrangle$}{$\llangle verbalization\rrangle$}|.

\begin{newpart}{MK: more definitions}
  In \sTeX, objects and concepts are represented by \textbf{symbols}, which are identified
globally and unambiguously by a URI that includes a \textbf{namespace}, a \textbf{module
  name}, and a \textbf{symbol name}. Every symbol can have one or more
\textbf{verbalizations}:
words of the domain jargon used to refer to the concept or object.
Verbalizations can vary between between subdomains and communities of practice. In
our example, the verbalization ``terminal [state]'' was annotated with the symbol with
name \lstinline|goal state|. It comes from the module with name \lstinline|search-problem|
in the namespace \lstinline|smglom/search/mod| imported via the \lstinline|\importmodule|
macro in line 1. As the symbol name \lstinline|goal state| is unique in the imports here
it suffices as a relative URI in the \lstinline|\sr| annotation.

If the symbol name and the desired verbalization coincide (as they often do in English),
the short-hand macro \lstinline|\sn| can be used (e.g. as \lstinline|\sn{utility}| for ``utility'' in \cref{lst:losa}).\ednote{
    FS: Do we need the last paragraph?
        I've removed the ``sn utility'' annotation from the listing
        because I thought it breaks the flow a bit.
        But if we want to show the sn macro, I can add it back.
}
\end{newpart}

@@

\begin{figure}[ht]\centering
  \fbox{\includegraphics[width=12cm]{../img/search-problem.en.pdf}}
\begin{lstlisting}[morekeywords={definame,symdecl,definiendum},numbers=left,
escapechar=!]
\begin{smodule}[title=Search Problem]{search-problem}
[... some imports ...]
\symdecl*{goal state}
\begin{sdefinition}
  [...] Certain \sns{state} are [...] \definame[post=s]{goal state} [...]
  (also called \definiendum{goal state}{terminal states}).
\end{sdefinition}
\end{smodule}
\end{lstlisting}
  \caption{Simplified definition of ``goal state'' from the domain model}\label{fig:state-space}
\end{figure}

\begin{newpart}{MK: trying to make some general remark about annotation before we go into
    semantic authoring}
    \subsection{Annotation Workflows and Requirements}\label{sec:workflows}

To annotate the word ``terminal'' in the document, the author has to be aware that it is a
technical term, that the module from \cref{fig:state-space} exists, know the symbol URI,
and manage redundancy of the imports -- i.e. only adding the \lstinline|\importmodule|
directive if it is not already (recursively) implied and possibly removing directives that
become redundant by the new one.  At ca 150,000 words in a semester's worth of lecture
notes and ca. 10--15\% of technical terms a rather daunting task.

@@

The most important resource for annotators is a verbalization to symbol URI mapping -- we
call it the \textbf{annotation catalog} -- either implicitly in the mind of the annotator
or explicitly as a technical artifact, which allows to
\begin{inparaenum}[\em i\rm)]
\item identify the known technical terms,
\item to map them to the corresponding symbols, and 
\item manage the necessary import directives.
\end{inparaenum}
This catalog can be harvested from the \textbf{formal declarations}
(\lstinline|\definiendum|, \lstinline|\sr|, \lstinline[language={}]|\begin|/\lstinline[language={}]|\end{smodule}| and
\lstinline|\importmodule|, etc.) of the domain model and existing documents in the
semantic corpus.

@@

But still, $L$-annotation catalogs are usually inadequate for $L$-annotation tasks in
practice, as
\begin{enumerate}[\em i\rm)]
\item $L$-verbalizations and symbol names do not often coincide and
\item the domain model may not even have $L$-modules -- especially for the ``smaller''
  languages
\end{enumerate}
 In this situation, we usually have to invest in
establishing a suitable catalog e.g. by providing $L$-translations (or at least dictionary
information) in the domain model or delineating modules and annotating definienda that the
domain model does not cover in the documents to be annotated.
\end{newpart}
\ednote{FS: Todo: verbalizations can also be
inferred from existing annotations (without any $L$-modules).}

@@

\section{Tool Support for Semantic Authoring in \sTeX/\ALeA}\label{sec:tools}
Semantic authoring can ``inherit'' tool support from both informal and formal authoring.
In the case of \sTeX, we can re-use spell checkers for traditional informal authoring.
There also is prototypical support for authoring \ALeA content in Microsoft Word
with a custom plugin (see \cite{KohKoh:woide24}),
which allows authors to use Word's spell checker and grammar checker.

On the formal side, there is a custom VSCode IDE plugin for \sTeX \cite{sTeX-IDE:git} that analyzes
annotations on the fly, displays the underlying reference semantics, reports errors and
redundancies, and even offers a concept search interface.
But such features only support
existing annotations and are, therefore, of limited use when creating new annotations,
which arguably is the most time consuming aspects of semantic authoring.
% \ednote{also: SPARQL API}

@@

Efficient support for the annotation of technical terms in natural language
has been elusive so far.

\section{The \snify System}\label{sec:snify}
The \snify\footnote{
From ``\texttt{\textbackslash sn}-ify''; ``\texttt{\textbackslash sn}'' is
a frequently used \sTeX short-hand macro for term annotations.}
utility \cite{stextools:git}
is a simple command-line tool that creates a catalog of
symbol-verbalization pairs, analyzes \sTeX source files, and then steps through all word
occurrences in the document that match a verbalization in the catalog.  Matching is based
on word stems using off-the-shelf stemmers, which allows us to e.g.\ match ``utilities''
with ``utility''.  For each matched word, the user is presented with an annotation choice
and interactions that allow to fine-tune the wordwise annotation
workflow.
The interface is inspired by traditional spell checkers like \lstinline|ispell|.
\Cref{fig:snify} shows a typical situation.

@@

As a companion to \snify,
we have \defianno, a prototype tool for annotating definienda in documents
that were originally authored without semantic annotations.
It can be configured to use different \LaTeX\ macros
to identify definienda (e.g.\ \lstinline|\emph| or \lstinline|\textbf|),
and then steps through them similarly to \snify.
For each potential definiendum, the user can then use fuzzy search to find
the corresponding symbol in the domain model.
This populates the catalog with verbalizations relevant to the document,
making subsequent \snify runs more effective,
especially for documents in languages that are not well covered by the
domain model (i.e.\ not English).

\section{Practical Evaluation and Future Work}\label{sec:eval}
\begin{wrapfigure}{r}{0.4\textwidth}
  \centering
  \includegraphics[width=0.4\textwidth]{../img/annocounts.pdf}
  \caption{Number of annotated terms in the \sTeX/\ALeA commons over time.}\label{fig:annocounts}
\end{wrapfigure}
In our experience, \snify's step-through workflow for annotating term references for symbols
from the domain model is almost an order of magnitude more efficient than writing the
annotations and imports by hand, even for annotators who are familiar with the domain
model.
Given the sheer number of term annotations (see \cref{fig:annocounts}),
this results in a considerable productivity gain.
For annotators unfamiliar with the domain model, the unassisted annotation task is
almost infeasible, and the average unfamiliarity naturally grows with the domain
model. The symbol disambiguation process -- on average a word induces 4.1 choices
% \ednote{
%     more details? median is 3, for existing annotations, the average is 2.6 and the median 2...
% }
-- is still manageable,
requires considerable concentration and domain knowledge, but little knowledge of the
domain model flexiformalization.

@@

For less complete domain models we have to skip too many terms that should ultimately be
annotated and annotation efficiency suffers.\ednote{
    FS: Isn't this wrong? We don't have to skip them, the problem is that \snify skips them.
    Efficiency does not suffer, but rather effectiveness.
}
This is currently the case for all
non-English languages in the \sTeX corpus we work with. Coverage in German is only about
half of that of English and we can already see the practical effects.
The \defianno tool (see end of \cref{sec:snify}) can mitigate this partially by
providing verbalizations for the definienda in the document,
but we also want to annotate terms that are not explicitly defined in the document.
A potential solution would be to extend \snify with a machine learning-based
tool for identifying unannotated technical terms that are not in the catalog.
The user could then pick a symbol for annotation, which would add a verbalization to the catalog.
In an earlier (pre-\snify) attempt, we used named-entity
recognition (NER) for classification of ``likely annotation candidate words''
\cite{hutterer:msc23}. However, this classification was not precise enough in the
distinction in ``technical terms'' and ordinary English noun phrases
to be practical on its own.
A combination with \snify's catalog-based approach might change the trade-offs involved.

% For less complete domain models we have to skip too many terms that should ultimately be
% annotated and annotation efficiency suffers. This is currently the case for all
% non-English languages in the \sTeX corpus we work with. Coverage in German is only about
% half of that of English and we can already see the practical effects. We have also
% experimented with a Slovene introductory math book, and it seems clear that apart from
% having to introduce a Slovene stemmer\ednote{MK@FS: there is one for python at
%   \url{https://repo.ijs.si/pboskoski/slo_stemmer}}, we would need to manually annotate all
% definienda in the book before we can harvest the symbol/verbalization pairs which are a
% prerequisite for annotation.\ednote{this is in principle possible with \defianno}
% \ednote{don't even need definienda - sn annos are enough to populate the catalog}
% \ednote{TODO: I've commented out Hutter in conclusion - we can integrate it here as a potential workflow}

One practical problem with \snify that remains is that the system only harvests
symbol/verbalization pairs from \sTeX files on the local file system. This means that, for
effective annotation, users (have to) download the whole content commons and keep it
updated. Unfortunately, the current MMT-based MathHub knowledge management
API\ednote{intro this above} does not supply verbalization information. The iMMT-based
API~\cite{iMMT:on} currently under development will, so \snify will be able to request the
pairs for the remaining (non-local) content commons.\ednote{
    This is already outdated.
    What is the current API status?
    How is it related to FLAMS?
}
If a pair is chosen in an annotation,
\snify can then automatically download the corresponding archive so that the annotated
file can be formatted with pdflatex or rusTeX locally.

@@

% In an earlier attempt to support semantic annotation in IDEs, we tried named-entity
% recognition (NER) for classification of ``likely annotation candidate words''
% \cite{hutterer:msc23}, however this classification was not precise enough in the
% distinction in ``technical terms'' and ordinary English noun phrases and named entities --
% the relevant task for annotation, and so made it impractical, especially, since it could
% not enter the actual annotations and import directives automatically. But maybe using the
% NER-based approach, after the \snify annotation process might change the trade-offs
% involved. In any case, the overall workflow suggested by the NER-based approach --
% especially after the verbalizations covered by the existing domain model\ednote{introduce
%   in the definition above that the domain model is all definitions, not only in the
%   smglom?} -- is geared towards adding modules and symbols -- the ones discovered by NER
% -- to the domain model.
